Automatic Matching of Bullet Lands

### 2017-01-12 imaging-paper.Rnw
Paper edits from Dr. Carriquiry

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is one of the primary goals. We propose an automated framework for the analysis of the 3D surface measurements, that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands (the surface between two bullet grooves) with too much damage to be suitable for comparison, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Rifling, manufacturing defects, and impurities in a barrel create striation marks on the bullet during the firing process. These marks are assumed to be unique to the barrel, as described in a 1992 AFTE article~\citep{afte:1992}. ``The theory of identification as it pertains to the comparison of toolmarks enables opinions of common origin to be made when the \emph{unique surface contours} of two toolmarks are in sufficient agreement" The article goes on to state that ``Significance is determined by the comparative examination of two or more sets of surface contour patterns comprised of individual peaks, ridges and furrows."

@@

Current standard practice~\cite{afte:1992} relies in part on the assessment of the so-called maximal number of consecutively matching striae (CMS), first defined by \citet{biasotti:1959}.
One of the problems is that a human inspection to determine CMS is subjective \citep{miller:1998}.

Here, we focus on the question of defining what constitutes a match, since there is no agreement in the forensics community. We propose a framework which allows for the automatic analysis of the surface topologies of bullets, and the transcription of the individual characteristics into a 2D plotting framework. 
This allows for an objective and quantitative assessment of striae-based bullet matches.

@@

Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in forensic science. An open-source package for working with x3p files has been written in R \citep{R} by Dr. Nicholas Petraco which we build upon in the development of our algorithm \citep{x3pr}. This package enables the direct reading and manipulation of x3p files.

@@

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". In the Hamby data, typical length (groove-to-groove) of a land is about \Sexpr{typical_width} micrometers or \Sexpr{typical_width / 1000} millimeters. For notational simplicity, we refer to a particular land of some bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the quality of a match. While we assess the performance of the algorithm in terms of a binary decision of match vs. non-match using a 50\% probability cut-off, our primary goal is to highlight the features that are statistically associated with matches and non-matches, and to provide a quantitative assessment of this association. In a real-world application of our algorithm, the raw scores would need further analysis and scrutiny, and it is likely a 50\% cut-off would be an inappropriate choice on the basis of reasonable doubt.

Our algorithm is fully open source and available on GitHub \citep{x3prplus}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, examine the factors that separate known bullet matches from non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.
The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods; aside from a zero false positive rate, no further details on a false-negative error rate for bullets are given nor are error rates for land-to-land matches mentioned. 
% Hamby study: automated systems were successfully used for matching:
%  Intelligent Automation's SciClops - Dr. Ben Bachrach (Maryland, United States);
% Automated Land Identification System (ALIS) - Mr. Tsuneo Uchiyama (Tokyo, Japan);
% Integrated Ballistics Identification System (IBIS) – Mr. Robert Thompson (California, United States);
% BulletTRAX-3DTM - Forensic Technology Scientists (Montreal, Canada) (2 sets)

Work has been done to quantify toolmark similarity by \citet{lock2013significance}. This algorithm determines an optimal matching window between two toolmark signatures, and then performs a set of both coordinated and independent shifts. Given a match, the coordinated shifts would be expected to yield correlation values higher than those obtained from independent shifts. This is assessed using a Mann-Whitney U Statistic. A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. This study is based on another set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

Modern automated techniques using 3D images have also been conducted by \citet{riva:2014}, but this work applies to striae found on cartridge cases and not bullets. This algorithm performs alignment of striae by rotation of the XY plane, which is not generalizable to bullets in which the XY plane is not flat. Other work on 3D images has been conducted by \citet{petraco:2012}, which also describes work carried out on cartridge cases, as well as screwdriver striation patterns, with promising results. We aim to use similar ideas of automated 3D image analysis, and apply this work to bullet surfaces with the modern tools enabled by the R statistical software.

The remainder of this paper is structured as follows: We first discuss two methods of modeling the class structure of the bullet surfaces. %We then describe a web application for visualizing and working with bullet surfaces. 
We then proceed to describing an automatic matching routine which we evaluate on the bullets made available through the Hamby study.

@@

In an approach to analyze the striation pattern, we extract a \emph{bullet profile} \citep{ma:2004} by taking a cross section of the surface measurements at a fixed height $x$ along the bullet land.
%An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the surface measurements across. 
Figure \ref{fig:fixedX} shows a plot of the side profile of a bullet land. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of a circular fit and the remaining residual structure can be found in Supplement section~\ref{supp:cylindrical}.
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

\subsection{Identifying groove locations}
%In an attempt to improve upon the previous algorithm, w
We first have to identify the locations of the left and right groove in the image. The grooves are assumed to contain no information relevant for determining matches. They also dominate the structure, and therefore need to be removed.  
Fortunately, the location and appearance of the grooves in the surface profiles is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, we can then follow the descent of the surface measurements inwards to the valley of the groove. 
The location of the valleys mark the points at which we should trim the image. The procedure can be described as follows:

@@

The smoothing factor $s$ introduced in the algorithm represents the window size to use for a rolling average. Higher values of $s$ therefore lead to more smoothed results. In the groove detection portion of this algorithm, we wish to remove most of the deviation but still maintain enough signal in the groove to detect it. Empirically, a value of $s = 35$ for the smoothing factor seems to work quite well (the smoothing factor is further discussed in Section~\ref{sec:smoothing}). It is important to note that the smoothing pass is done twice. That is, the smoothed data is once again smoothed by computing a new rolling average with the same smoothing factor. This bears some similarities to the ideas of John Tukey in his book Exploratory Data Analysis, where he describes a smoothing process called ``twicing" in which a second pass is made on the residuals computed from the first pass and then added back to the result~\citep{tukey:1977}. This has the effect of introducing a bit more variance back into the smoothed data. 
We instead performed a second smoothing pass on the smoothed data, which has the effect of weighting observations near the center of the window the highest, with the weights linearly dropping off as we reach either end of the smoothing window.

@@

The main idea of locally weighted regression is to use a weighting routine that  emphasizes the effect of points close by and de-emphasizes the effect of points as they are further away. The weighting function used here is the tricubic function $w(d) = \left(1 - d^3\right)^3$, for $d \in [0,1]$ and $w(d) = 0$ otherwise. Here, $d$ is defined as the distance between $x_i$ and the location of the fit $x_o$ and the maximum distance of the range of the $x$-values for span $\alpha$ in $x_o$.

@@

\section{Automatic matching}
Applying the loess fit to a range of different signatures (see Figure~\ref{fig:manualmatch-rgl} for signatures extracted at heights between 50$\mu m$ and 150~$\mu m$) shows the 3D striation marks from two bullets. Signatures of bullet~1 are shown on the left (all extracted from heights below 100$\mu m$) and signatures of bullet 2 are shown on the right (extracted at heights above 100$\mu m$). Signatures are manually aligned, resulting in many of the striation marks to continuously pass from one side to the other. Visually, this allows for an easy assessment of these two bullet lands as a match. However, this match relies on visual inspection and can therefore not be completely objective.  The goal of this section is to eliminate the need for a visual inspection during the matching process and replace it by an automatic algorithm. This also allows for a quantification of the quality of the match.

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bulletname, linetype=bulletname)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Signatures (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet land and extract the signature at the lowest height in this region, because typically, individual characteristics are best expressed at the lower end of the bullet (see Supplement section~\ref{supp:bulletbottom} for a more detailed discussion).

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} using a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} of the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify joint intervals between two signatures as the areas in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features could be extracted. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item The value of the cross-correlation function (ccf) between the aligned signatures \cite{vorburger:2011},
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

Counting the maximal number of CMS is part of the current practice to identify bullet matches~\citep{nichols:1997, nichols:2003, nichols:2003b}. 
In the example of Figure~\ref{fig:match}, the number of consecutive matching striations (CMS) is fifteen, a high number indicative of a match between the bullets. 
 Note that the definition of CMS given in this paper does not match the one given in \citet{thompson:2013}. There, CMS is defined only in terms of matching peaks without regarding valleys. Additionally, peaks in  \citet{thompson:2013} are  used only if they can be identified and matched `within a tolerable range' between lands. The definition given here is computationally less complex, but should yield highly correlated values, because of the requirement to only consider signatures from a stable region in the land (see Section~\ref{sec:heights} for further details on stability of regions). In the Hamby study, the definition of CMS by \citet{thompson:2013} leads to approximately half of the values of CMS defined in this paper (with a correlation coefficient between the values of the two definitions of about 0.92). 
For lead bullets, such as used in the Hamby study, \citet{biasotti:1959} considered four or more consecutive peaks (corresponding to eight or more consecutive lines in our definition) to be sufficient evidence of a match.  %We will assess the claim that eight or more consecutive lines are sufficient for a match in the next section.

@@

\subsection{Impact of bullet height}\label{sec:heights}
The height at which signatures are extracted for a comparison between bullet lands matters -- signatures taken from heights that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity for quality control: we have to be aware of the height that was used in a matching. Visually, matches degrade if the signatures upon which the match is based are from heights further than 200$\mu m$ apart  (see Supplement section~\ref{supp:ccf} for a more thorough discussion). However, we can extract signatures from multiple heights of the same bullet land for an initial assessment of its quality. By comparing signatures from heights that are not too far apart --  25~$\mu m$ to 50~$\mu m$ -- we get an indication whether the signatures come from a rapidly changing section of the surface, indicative of a break-off or some other damage, or from a stable section, where we have a reasonable expectation of finding matches to other signatures. In the approach here, we keep increasing the height $x$ at which the signature is taken until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet~1-1 from barrel 3, where `stability' is  defined as two aligned signatures from heights chosen 25$\mu m$ apart having a cross-correlation of at least 0.95.

@@

Figure~\ref{fig:varysmooth} displays the  peaks and valleys detected in the same signature at smoothing factors of 5, 25, and 45, respectively. The dark line corresponds to the smoothed values, while the grey line in the back shows the raw signature. The choice of smoothing factor is a classical decision of a bias/variance trade-off. It is immediately clear that a small smoothing factor like 5 is a poor choice. It results in a significant amount of noise in the data such that even just a point or two can skew the rolling average enough for a peak or valley to be detected. Given that striation widths are typically much larger, we are in effect muddying the waters by performing such minimal smoothing. Another consideration is that the smoothing should not fall below the  resolution of the equipment at which the surface measurements are taken -- in order to not introduce artifacts in the analysis.

>>>>

\address{Address of the Authors\\
Department of Statistics and Statistical Laboratory\\ Snedecor Hall\\
Ames Iowa 50011-1210\\
\printead{e1}\\
\phantom{E-mail:\ }\printead*{e2}\\
\phantom{E-mail:\ }\printead*{e3}}
\end{aug}
%
\begin{abstract}
In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. Firearm examination is a forensic tool used to help the court determine whether two bullets were fired from the same gun barrel. During the firing process, rifling, manufacturing defects, and impurities in the barrel create striation marks on the bullet. Identifying these striation markings in an attempt to match two bullets is one of the primary goals of firearm examination. We propose an automated framework for the analysis of the 3D surface measurements of bullet lands that first transcribes the markings into a 2D plotting framework. This makes identification of matches easier and allows for a quantification of both matches and matchability of barrels. The automatic matching routine we propose manages to (a) correctly identify lands (the surface between two bullet grooves) with too much damage to be suitable for comparison, and (b) correctly identify all 10,384 land-to-land matches of the James Hamby study \citep{hamby:2009}.
\end{abstract}

@@

Rifling, manufacturing defects, and impurities in a barrel create striation marks on the bullet during the firing process. These marks are assumed to be unique to the barrel, as described in a 1992 AFTE article~\citep{afte:1992}. ``The theory of identification as it pertains to the comparison of toolmarks enables opinions of common origin to be made when the \emph{unique surface contours} of two toolmarks are in sufficient agreement". The article goes on to state that ``Significance is determined by the comparative examination of two or more sets of surface contour patterns comprised of individual peaks, ridges and furrows."

@@

Current firearm examination practice relies mostly on visual assessment and comparison of striation. Indeed, the AFTE Theory of Identification (https://afte.org/about-us/what-is-afte/afte-theory-of-identification) explicitly requires that examiners evaluate the strength of similarity between two samples relative to other comparisons they may have carried out in the past. An attempt to quantify the degree of similarity consists in counting the number of consecutively matching striae (CMS) between two bullets, first proposed by \citet{biasotti:1959}. This approach has two drawbacks, however. First, determining matching striae is still a subjective activity. Second, as discussed by \citet{miller:1998}, the number of CMS may be high even if the bullets were not fired by the same gun.

Here, we focus on the question of defining a metric that can be used to objectively compare two bullets. We propose a framework which allows for the automatic analysis of the surface topologies of bullets, and the transcription of the individual characteristics into a 2D plotting framework. 
This allows for an objective and quantitative assessment of striae-based bullet matches.

@@

We work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce 20 known test bullets and 15 unknown bullets for comparison. 
3D topographical images of each bullet were obtained using a NanoFocus lens at 20x magnification and made publicly available on the NIST Ballistics Database Project\footnote{\url{http://www.nist.gov/forensics/ballisticsdb/hamby-consecutively-rifled-barrels.cfm}} in a format called x3p (XML 3-D Surface Profile). The x3p format conforms to the ISO5436-2 standard\footnote{\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}}, implemented to provide a simple and standard conforming way to exchange 2D and 3D profile data. It was adopted by the OpenFMC (Open Forensic Metrology Consortium\footnote{\url{http://www.openfmc.org/}}), a group of academic, industry, and government firearm forensics researchers whose aim is to establish best practices for researchers using metrology in forensic science. We have developed an open-source package for analyzing bullet lands written in R \citep{R}. This package is called bulletr \citep{bulletr} and enables direct reading and manipulation of x3p files. It also implements all of the methods we propose in this paper. A different package exists for reading x3p files called x3pr \citep{x3pr}, developed by Petraco (2014), but it is not designed to carry out calculations like the ones we propose after the x3p files have been read.

@@

Each fired bullet is provided in the form of a set of six x3p files, where each file is a surface scan between adjacent grooves on the bullet, called a ``land". In the Hamby data, typical length (groove-to-groove) of a land is about \Sexpr{typical_width} micrometers or \Sexpr{typical_width / 1000} millimeters. For notational simplicity, we refer to a particular land of a bullet as bullet X-Y, where X is the bullet identifier, and Y is the land number. An example of plotting one of these lands is given in Figures~\ref{fig:sidex3p} and~\ref{fig:topx3p}. These figures show side and top profiles of the land respectively. The tilt of the lines to the left in Figure~\ref{fig:topx3p} is not an artifact, but a direct and expected consequence of the spin induced by the rifling during the firing process. Depending on whether a barrel is rifled clockwise or counter-clockwise, the striations have a left or right tilt. The direction of the rifling is a class characteristic, i.e.\ a feature that pertains to a particular class of firearms, and is not unique at the individual barrel/bullet level.

@@

The purpose of our paper is to present an automatic matching routine that allows for a completely objective assessment of the strength of a match between two bullet lands. While we assess the performance of the algorithm in terms of a binary decision of match vs. non-match using a 50\% probability cut-off, our primary goal is to highlight the features that are statistically associated with matches and non-matches, and to provide a quantitative assessment of this association. In a real-world application of our algorithm, the raw scores would need further analysis and scrutiny, and it is likely a 50\% cut-off would be an inappropriate choice on the basis of reasonable doubt.

Our algorithm is fully open source and available on GitHub \citep{bulletr}. This transparency allows for a greater understanding of the individual steps involved in the bullet matching process, and allows other forensic examiners, as well as outside observers, to examine the factors that discriminate between known bullet matches and non-matches. We have chosen to perform the matching on a land-to-land level, rather than bullet-to-bullet level. Although doing so introduces an implicit assumption of independence between lands, assuming independence only serves to make the task more challenging.

The remainder of this paper is structured as follows: We first briefly review some earlier work. We then discuss two methods of modeling the class structure of the bullet surfaces. Finally, we proceed to describing an automatic matching routine which we evaluate on the bullets made available through the Hamby study.

\section{Previous work}

There have been attempts to develop automatic or semi-automatic matching protocols, but most have focused on breech face and firing pin marks \citep[e.g.][]{riva:2014} or discuss a single attribute for comparison \citep[e.g.][]{vorburger:2011, chu:2011}. Still others refer to proprietary algorithms \citep{roberge:2006}. We briefly review some of this earlier work in what follows.

The original paper on the complete Hamby study already reports the successful use of several computer-assisted methods. However, aside from a zero false positive rate, false-negative error rates for bullets are not given nor are error rates for land-to-land matches mentioned.

\citet{lock2013significance} proposed an approach to quantify similarity of toolmarks. Their algorithm determines an optimal matching window between two toolmark signatures, and then performs a set of both coordinated and independent shifts. Given a match, the coordinated shifts would be expected to yield correlation values higher than those obtained from independent shifts. This is assessed using a Mann-Whitney U Statistic.

A procedure for bullet matching using the BulletTrax3D system is described in \citet{roberge:2006}. Their study used a different set of ten consecutively rifled barrels; matches are identified based on a bullet-to-bullet correlation score. The authors state that this process `could be automated', but no implementation of the algorithm is available.

Modern automated techniques using 3D images have also been proposed by \citet{riva:2014}. However, the authors focused on cartridge cases and not bullets. This might seem like a trivial distinction, but it has implications for the development of the algorithm. The algorithm performs alignment of striae by rotation of the XY plane, which is not generalizable to bullets in which the XY plane is not flat.

Other work on 3D images has been described by \citet{petraco:2012}, who also focus on cartridge cases, as well as screwdriver striation patterns, and by others \citep[e.g.,][]{chu:2010, chu:2011, vorburger:2011}.

@@

To analyze the striation pattern, we extract a \emph{bullet profile} \citep{ma:2004} by taking a cross section of the surface measurements at a fixed height $x$ along the bullet land.
%An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the surface measurements across. 
Figure \ref{fig:fixedX} shows a plot of the side profile of a bullet land. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of a circular fit and the remaining residual structure can be found in Supplement Section~\ref{supp:cylindrical}.
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

@@

\subsection{Identifying groove locations}
%In an attempt to improve upon the previous algorithm, w
We first identify the location of the left and right grooves in the image. The grooves are assumed to contain no information relevant for determining matches. They also dominate the structure, and therefore need to be removed.  
Fortunately, the location and appearance of the grooves in the surface profiles is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, and we can then follow the descent of the surface measurements inwards to the valley of the groove. 
The location of the valleys mark the points at which we trim the image. The procedure can be described as follows:

@@

The smoothing factor $s$ introduced in the algorithm represents the window size to use for a rolling average. Higher values of $s$ therefore lead to more smoothing. Empirically, a value of $s = 35$ for the smoothing factor seems to work well (the smoothing factor is further discussed in Section~\ref{sec:smoothing}). It is important to note that the smoothing pass is done twice. That is, the smoothed data are once again smoothed by computing a new rolling average with the same smoothing factor. This bears some similarities to the ideas of John Tukey in his book Exploratory Data Analysis, where he describes a smoothing process called ``twicing" in which a second pass is made on the residuals computed from the first pass and then added back to the result~\citep{tukey:1977}. This has the effect of introducing a bit more variance back into the smoothed data. 
We instead performed a second smoothing pass on the smoothed data, which has the effect of weighting observations near the center of the window the highest, with the weights linearly dropping off as we reach either end of the smoothing window.

@@

The main idea of locally weighted regression is to use a weighting routine that  emphasizes the effect of points close to the fitting location and de-emphasizes the effect of points as they are further away. The weighting function used here is the tricubic function $w(d) = \left(1 - d^3\right)^3$, for $d \in [0,1]$ and $w(d) = 0$ otherwise. Here, $d$ is defined as the distance between $x_i$ and the location of the fit $x_o$ and the maximum distance of the range of the $x$-values for span $\alpha$ in $x_o$.

@@

\section{Automatic matching}
Applying the loess fit to a range of different signatures (see Figure~\ref{fig:manualmatch-rgl} for signatures extracted at heights between 50$\mu m$ and 150~$\mu m$) shows the 3D striation marks from two bullets. Signatures of bullet~1 are shown on the left (all extracted from heights below 100$\mu m$) and signatures of bullet 2 are shown on the right (extracted at heights above 100$\mu m$). Signatures are manually aligned, resulting in many of the striation marks to continuously pass from one side to the other. Visually, this allows for an easy assessment of these two bullet lands as a match. However, this match relies on visual inspection and is therefore subjective.  The goal of this section is to eliminate the need for a visual inspection during the matching process and replace it by an automatic algorithm. This also allows for a quantification of the strength of the match.

@@

ggplot() + 
  geom_rect(aes(xmin = xmin, xmax = xmax, fill=factor(type)), ymin = -6, ymax=6.5,  data = lines, alpha=0.2, show.legend = FALSE) +
  theme(legend.position="bottom") +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(lines, match)) +
  ylim(c(-6,6.5)) + theme_bw() +
  geom_line(data=lofX, aes(x=y, y=l30, group=bulletname, linetype=bulletname)) +
  scale_linetype_discrete("") +
  scale_colour_manual("", values=cols) +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey60", 0.5)) +
    theme(legend.position = c(1,1.1), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab(expression(paste("Signatures (in ",mu,"m)", sep=""))) +
  xlab("y")
@
    }
\end{subfigure}
\caption{\label{fig:match}Matching striation marks: smooth (a), identify peaks and valley (b), and match peaks and valleys between signatures (c).}
\end{figure}
%\afterpage{\clearpage}
Figure~\ref{fig:match} gives an overview of the automated matching routine: 
We first identify a stable region for each bullet land and extract the signature at the lowest height in this region, because typically, individual characteristics are best expressed at the lower end of the bullet (see Supplement Section~\ref{supp:bulletbottom} for a more detailed discussion).

All of the other steps are done on pairs of bullet lands:
\begin{enumerate}
\item {\bf Smooth the two signatures} using a loess with a very small span (see Figure~\ref{fig:smooth}). 
\item Use cross-correlation to {\bf find the best alignment} of the two signatures: shift one of the signatures by the lag indicated by the cross-correlation function (see Figure~\ref{fig:ccf} for the cross-correlation function and Figure~\ref{fig:crosscutX} for the resulting shift).
\item Using a rolling average, {\bf identify peaks and valleys} for each of the signatures. We then define an interval around the location of the extrema on each side as one third of the distance to the location of the next extrema (see Figure~\ref{fig:smoothcutb}). Peaks and valleys constitute the \emph{striation marks} on the bullet.
\item {\bf Match striations across signatures:} based on the intervals around the extrema as defined above, we identify common intervals as the areas in which two or more of the individual intervals overlap: a joint interval is defined as the smallest interval that encompasses all of the overlapping intervals. A joint interval is then called a match(ing stria) between the signatures, if all of the intervals are of the same type of extrema, i.e.\ they are either all peaks or all valleys. In Figure~\ref{fig:match} all matches are shown as color-filled rectangles corresponding to their type of extrema (peaks are shown in orange, and valleys in green). Non-matching intervals are left grey. 
\item {\bf Extract features from the aligned signatures and the matches between them:} many different features can be extracted from the aligned signatures. Here, we describe a few of the ones that can be found in the literature and some that we found to be of practical relevance:
\begin{enumerate}[label=(\roman*)]
\item Maximal number of CMS (consecutive matching striae), and, similarly, the number of consecutively non-matching striae (CNMS), 
\item Number of matches and non-matches,
\item The value of the cross-correlation function (ccf) between the aligned signatures \citep{vorburger:2011},
\item Average difference $D$ between signatures, defined as the Euclidean vertical distance between surface measurements of aligned signatures. Let $f(t)$ and $g(t)$ be smoothed, aligned signatures:
\[
D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2,
\]
\item The sum $S$ of average absolute heights of matched extrema: for each of the two matched stria, compute the average of the absolute heights of the peaks or valleys. $S$ is then defined as the sum of all these averages. 
\end{enumerate}
\end{enumerate}
%
The difference $D$ between signatures is here defined as the Euclidean distance (in $\mu m$). In the paper by \citet{ma:2004}, distance is defined as a measure relative to the first signature, which serves as a comparison reference and is therefore a unitless quantity.

Counting the maximal number of CMS is part of the current practice to identify bullet matches~\citep{nichols:1997, nichols:2003, nichols:2003b}. 
In the example of Figure~\ref{fig:match}, the number of consecutive matching striations (CMS) is fifteen, a high number suggestive of a match between the lands. Note that the definition of CMS we use does not match the one given in \citet{thompson:2013}. There, CMS is defined only in terms of matching peaks without regarding valleys. Additionally, peaks in  \citet{thompson:2013} are  used only if they can be identified and matched `within a tolerable range' between lands. The definition given here is computationally less complex, but should yield highly correlated values, because of the requirement to only consider signatures from a stable region in the land (see Section~\ref{sec:heights} for further details on stability of regions). In the Hamby study, the definition of CMS by \citet{thompson:2013} leads to approximately half of the values of CMS defined in this paper (with a correlation coefficient between the values of the two definitions of about 0.92). 
For lead bullets, such as used in the Hamby study, \citet{biasotti:1959} considered four or more consecutive peaks (corresponding to eight or more consecutive lines in our definition) to be sufficient evidence of a match.  %We will assess the claim that eight or more consecutive lines are sufficient for a match in the next section.

@@

\subsection{Impact of bullet height}\label{sec:heights}
The height at which signatures are extracted for a comparison between bullet lands matters -- signatures taken from heights that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity for quality control: we have to be aware of the height that was used in a matching. Visually, matches degrade if the signatures upon which the match is based are from heights further than 200$\mu m$ apart  (see Supplement Section~\ref{supp:ccf} for more discussion). However, we can extract signatures from multiple heights of the same bullet land for an initial assessment of its quality. By comparing signatures from heights that are not too far apart --  25~$\mu m$ to 50~$\mu m$ -- we get an indication whether the signatures come from a rapidly changing section of the surface, indicative of a break-off or some other damage, or from a stable section, where we have a reasonable expectation of finding matches to other signatures. In the approach here, we keep increasing the height $x$ at which the signature is taken until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet~1-1 from barrel 3, where `stability' is  defined as two aligned signatures from heights chosen 25$\mu m$ apart having a cross-correlation of at least 0.95.

@@

Figure~\ref{fig:varysmooth} displays the  peaks and valleys detected in the same signature at smoothing factors of 5, 25, and 45, respectively. The dark line corresponds to the smoothed values, while the grey line in the back shows the raw signature. The choice of smoothing factor is a classical decision of a bias/variance trade-off. It is immediately clear that a small smoothing factor like 5 is a poor choice. It results in a significant amount of noise in the data such that even just a point or two can skew the rolling average enough for a peak or valley to be detected. Given that striation widths are typically much larger, we are in effect muddying the waters by performing such minimal smoothing. Another consideration is that the smoothing should not fall below the  resolution of the equipment at which the surface measurements are taken -- so as to not introduce artifacts in the analysis.
